Generative Modeling of Cell Images

### 2017-10-16 paper/paper.tex
Update to paper

% to compile a camera-ready version, add the [final] option, e.g.:
\usepackage[final]{nips_2017}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography

\title{CytoGAN: Synthesis and Representation Learning of Microscopy Cell Images}

\author{
  Peter Goldsborough \\
  Facebook, Inc.\\
  Menlo Park, CA, USA \\
  \texttt{psag@fb.com} \\
  \And
  Nick Pawlowski \\
  Imperial College London \\
  London, UK \\
  \texttt{n.pawlowski16@imperial.ac.uk} \\
  \AND
  Juan C. Caicedo \\
  Imaging Platform \\
  Broad Institute of MIT and Harvard \\
  Cambridge, MA, USA \\
  \texttt{jccaicedo@broadinstitute.org} \\
  \And
  Shantanu Singh \\
  Imaging Platform \\
  Broad Institute of MIT and Harvard \\
  Cambridge, MA, USA \\
  \texttt{shsingh@broadinstitute.org} \\
  \And
  Anne E Carpenter \\
  Imaging Platform \\
  Broad Institute of MIT and Harvard \\
  Cambridge, MA, USA \\
  \texttt{anne@broadinstitute.org} \\
}

@@

\begin{abstract}
  The abstract paragraph should be indented \nicefrac{1}{2}~inch
  (3~picas) on both the left- and right-hand margins. Use 10~point
  type, with a vertical spacing (leading) of 11~points.  The word
  \textbf{Abstract} must be centered, bold, and in point size 12. Two
  line spaces precede the abstract. The abstract must be limited to
  one paragraph.
\end{abstract}

\section{Introduction}

Advances in high-throughput microscopy systems have enabled acquisition of large volumes of high-resolution cell images, paving the way for novel computational methods that can leverage these large quantities of data, both in order to assist biologists with tasks such as morphological profiling, which aims to capture quantitative differences in the phenotypes of cellular perturbations, as well automate certain image analysis pipelines, like mechanism-of-action (MOA) prediction for drug compounds, in their entirety, end-to-end.

% The transition here feels unsmooth

Morphology-based profiling can be stated as the task of mapping microscopic cell images to representations that capture the most salient features and axes of variation of each cell's phenotype in an unsupervised manner, ideally partitioning the representation space cleanly into clusters of cells with similar visible properties and in the case of drug compounds, similar function. Two current methods that tackle this challenge successfully are specialized software like CellProfiler, which extracts representations via manually tuned features and traditional computer vision pipelines, as well as, more recently, transfer learning approaches using features learned by deep convolutional neural networks originally trained to classify miscellaneous objects.

Being an unsupervised learning problem, morphological profiling naturally invites investigation into the effectiveness of suitable approaches employing deep neural network architectures. Indeed, there have been attempts at capturing representations from cell images using autoencoders and deep variational methods, albeit with limited success. A recent addition to the unsupervised learner's toolbox is the adversarial framework introduced by Goodfellow et al. Generative adversarial networks (GANs) in particular have shown to excel at the task of synthesizing realistic natural images. In this work, we present findings on the effectiveness of GANs for two tasks in the domain of cell microscopy imaging: First, we show that GANs can generate highly realistic and biologically consistent cell images while providing greater intuition and interpretability over the synthesis mechanism than other generative methods. Second, we investigate ways to capture the internal representations of adversarial networks trained on fluorescence microscopy images of cultured human cells as a solution to morphological profiling and evaluate the quality of learned features for the task of predicting the mechanism-of-action of drug compounds.

@@

There exists a respectable body of work in the two lines of research that our work intersects: automated morphological profiling, and representation learning with deep neural networks, in particular generative architectures.

Caicedo et al. recently outlined the state and challenges of the morphological profiling problem in [], where the authors describe best practices for profiling pipelines using both traditional as well as deep learning methods. Prior to this, Ljosa et al. contrast the quality of representations composed of manually selected features with various augmentations and evaluate them at the task of MOA prediction on the BBBC021 dataset, consisting of microscopic images of human breast cancer cells (MCF7) treated with different compounds at differing concentrations. Both [Singh] and [Mike Ando] build on this work by showing how correcting illumination biases in images of the same dataset and applying statistical whitening transformations to learned representations, respectively, can ameliorate results reported by [Ljosa et al]. In [Nick], Pawlowski et al. for the first time report of a representation-learning method based on deep learning that is competitive with hand-picked features at the task of MOA prediction. In this work, feature layers from Inception-v3 networks trained on assorted consumer images are treated as cell representations, leveling prediction accuracy reported by [Singh].

Outside the domain of biological images, solutions to unsupervised representation learning have been proposed by [Hinton, Salakuthdinov] in the form of autoencoders and its many variations such as denoising autoencoders [], variational autoencoders [] and stacked autoencoders []. Transfer learning of convolutional network features as employed by [Nick] had first been investigated by [someone]. Preceding all of the above, Hinton et al. studied the effectiveness of restricted boltzman machines and deep belief networks at the representation learning task. More recently, the ladder network architecture was proposed by Rasmus to learn representations in a semi-supervised setting with promising results.

% Just delete this whole paragraph up there?

While both morphological profiling and unsupervised representation learning have been studied extensively in isolation, few works have synthesized the two. Nick et al. first investigated autoencoders and variational autoencoders to capture semantic representations of microscopic cell images in the BBBC021 dataset but reported results far inferior to hand-tuned features or transfer-learning approaches. The work most similar to ours is [], where Osokin et al. first study the task of cell image synthesis using generative adversarial networks and show that generated images are visually realistic and the latent space semantically smooth and biologically interpretable.

\section{Cell Image Synthesis}

In this section, we present our methodologies and findings for the task of generating realistic cell images with adversarial architectures. We begin by introducing the general GAN framework, outline the nature of the data we train on, and finally discuss our results as well as the qualitative and quantitative differences between a number of GAN models we tried.

Generative adversarial networks were first described in a seminal work by Goodfellow et al. [Foo]. The architecture they present consists of two \emph{players}, a \emph{generator} network $G$ and a \emph{discriminator} or \emph{critic} network $D$. The former receives samples $\mathbf{z} \sim P(\mathbf{z})$ drawn from a noise prior $P(z)$, typically a multi-variate uniform or normal distribution, which it maps to values $G(z)$ from the generator distribution $P_{model}$. The discriminator network must learn to distinguish such synthetic samples from real values $\mathbf{x}$ drawn from a data prior $P_{data}(\mathbf{x})$, with the goal of maximizing the distance between the generated distribution $P_{model}$ and data prior $P_{data}$. Feedback from the discriminator is in turn provided to the generator, which follows the opposite goal of fooling the discriminator into believing its synthesized samples stem from $P_{data}$. This formulation leads to a zero-sum game where the loss $L_G$ of the generator network is the negative of the discriminator's loss $L_D$, with the game's solution lying in a nash equilibrium found by playing the minimax game

$$\min_G \max_D V(G, D) = \mathbb{E}_{x \sim P_{data}}[\log(D(\mathbf{x})] + \mathbb{E}_{z \sim P_{noise}}[\log(1 - D(G(\mathbf{z}))]$$

% Trim this down ^

We evaluate the GAN framework at the task of synthesizing realistic and biologically consistent cell images. Our data prior $P_{data}$ consists of microscopy images from the BBBC021 dataset of MCF7 human breast cancer cell lines [] that were treated with 113 different drug molecules at 8 concentrations. Three \emph{channels} are available for each image in BBBC021, corresponding to DNA, F-Actin and B-Tubulin structures obtained from fluorescence microscopy. We stack these channels and treat them as RGB images via a simple $DNA \mapsto Red, Actin \mapsto Green, B-Tubulin \mapsto Blue$ mapping. Images in BBBC021 contain multiple cells and are not segmented. We segment them using the CellProfiler software, such that $P_{data}$ consists of segmented, single-cell images. Finally, we normalize the luminance of each channel to the range $[0-255]$ to effectively brighten the image, which improved the quality of generated images in our experience.

We evaluated two \emph{families} of GAN architectures. The first includes the Deep Convolutional GAN (DCGAN) model proposed by Radford et al., which models the generator network as a series of convolution and deconvolution (up-sampling?) operations that successively transform and up-sample the noise sample $z$ into an image $G(z)$ that ideally resembles those in $P_{data}$. The discriminator $D$ is a conventional deep convolutional neural network that maps images $x$ to a probability $D(x)$ representing the discriminator's confidence that the image stems from $P_{data}$. In general, the GAN game is notoriously unstable and difficult to optimize, as either player can overpower the other, leading to the discriminator giving the generator no chance to improve, or the generator exploiting only a single mode of the data distribution -- a phenomenon known as \emph{mode collapse} (this sentence needs to be better!!). We experienced this instability first-hand with the DCGAN model, with mode-collapse occurring frequently and very early on in the training process, leading to dissatisfying results. Derivations of the DCGAN such as the Least Squares (LSGAN) [cite] or Wasserstein (WGAN) [cite] architecture address the inherent instability of the GAN game and we were able to train both models to convergence. We report best qualitative results for the LSGAN model, which produced highly realistic and biologically consistent synthetic images, as shown in Figure 1.

\begin{figure}[h]
  \centering
  \fbox{\rule[-.5cm]{0cm}{4cm} \rule[-.5cm]{4cm}{0cm}}
  \caption{Sample figure caption.}
\end{figure}

% Need to elaborate why the images make sense biologically, like that the actin cradles the nucleus

The second family of models we successfully trained to convergence that we deem noteworthy is the class of Boundary Equilibrium GAN [cite] models proposed by [Foobar et al]. BEGAN replaces the discriminator in DCGAN with an autoencoder and uses the reconstruction loss of this autoencoder as a measure of whether the image stems from $P_{data}$ or $P_{model}$. % I think we have no room for BEGAN here?

Besides generating realistic cell images, we additionally found that interpolations between noise samples drawn from $P_{noise}$ were visually smooth. We draw two variables $z_1, z_2$ from the noise prior $P_{noise}$ and sample images for noise samples at various distances between $z_1$ and $z_2$. In all of our experiments, we used a multi-variate Gaussian distribution for the noise distribution, thus we use the spherical linear interpolation strategy from [cite] to transition between $z_1$ and $z_2$. Figure 2 shows that images change in a smooth fashion between the two points of the noise space.

\begin{figure}[h]
  \centering
  \fbox{\rule[-.5cm]{0cm}{4cm} \rule[-.5cm]{4cm}{0cm}}
  \caption{Sample figure caption.}
\end{figure}

Radford et al. originally reported on intriguing relationships found between regions of the noise distribution, where algebra on vectors in the noise space translated directly to corresponding semantic changes in generated images. This is similar to results studied by Mikolov et al. for word embeddings, where distances between words in the embedding space proved surprisingly rich and linear in nature. We present similar findings in Figure 3 (have to make).

% Can we add something about the biological relevance of all of this? Like why semantic relationships ^ are neat?

\section{Representation Learning}

In this section, we investigate the ability of GANs to infer vector representations of images for the purpose of profiling. We evaluate the quality of learned representations in both quantitative and qualitative experiments. First, we discuss the performance of GAN representations in the well-studied task of MOA prediction. Then, we show that representation learned by GANs have interesting semantic relationships.

The original GAN framework proposed by Goodfellow et al. does not incorporate an explicit means of inference. As such, we require extensions to traditional GAN architectures that allow us to map images drawn either from $P_{model}$ or $P_{data}$ to latent representations via some \emph{encoding} function $E(x)$. The simplest such extension is to add a hidden layer to the end of the discriminator network $D$ that induces a latent space. The intuition being that the discriminator must learn very strong internal representations in order to perform its task of separating synthetic from real images. Another benefit of this strategy is that it adds almost no computational overhead, since all but the final layer of the discriminator is shared with the encoder.  Since the BEGAN architecture models the discriminator as an autoencoder, it has a bottleneck layer that naturally provides latent representations. We compute MOA classification accuracy for LSGANs with an added hidden layer, as well as BEGANs, and report our findings in Table 1.

\begin{table}[t]
  \caption{Sample table title}
  \label{sample-table}
  \centering
  \begin{tabular}{lll}
    \toprule
    \multicolumn{2}{c}{Part}                   \\
    \cmidrule{1-2}
    Name     & Description     & Size ($\mu$m) \\
    \midrule
    Dendrite & Input terminal  & $\sim$100     \\
    Axon     & Output terminal & $\sim$10      \\
    Soma     & Cell body       & up to $10^6$  \\
    \bottomrule
  \end{tabular}
\end{table}

% Discussion?

We found that the latent representations learned by either of the two architectures did not outperform other current methods such as transfer learning or mean CellProfiler profiles. Even though MOA classification accuracy is not yet competitive, we believe that representation learning of cell images provides benefits over established methods:

\begin{itemize}
	\item Less manual tuning and faster than CellProfiler,
    \item More tuned to actual data since it is trained on cell images, providing more room for biological extensions,
    \item Results are better than for other generative models such as VAEs
\end{itemize}

@@

We investigate use cases of generative adversarial networks (GANs) for the domain of cell miscroscopy imaging. First, we show that GANs have the ability to synthesize highly realistic images of cells that respect natural phenomena and explain how this is biologically useful (NOT YET). Second, we break ground on the ability of GANs to learn representations of cell images and establish their performance at the task of drug compound mechanism-of-action prediction. We discuss that even though the representations learnt by adversarial models are currently inferior at this task, GANs provide valuable benefits over each of the competing approaches.

We emphasize that this work has shed light only on the tip of the iceberg of how the adversarial framework is useful for the field of computational biology. Due to the generality of the adversarial framework, there remain many more avenues to explore. For example, we hope that future work investigates recently proposed bidirectional GAN models or other, novel approaches to infering representations under the GAN umbrella. Finally, we note that deep neural network architectures such as those described in this work flourish with large amounts of data. As such, we conjecture that the performance of approaches outlined in this work will improve if and when larger biological image datasets become available.

\section*{References}

References follow the acknowledgments. Use unnumbered first-level
heading for the references. Any choice of citation style is acceptable
as long as you are consistent. It is permissible to reduce the font
size to \verb+small+ (9 point) when listing the references. {\bf
  Remember that you can go over 8 pages as long as the subsequent ones contain
  \emph{only} cited references.}
\medskip

\small

[1] Alexander, J.A.\ \& Mozer, M.C.\ (1995) Template-based algorithms
for connectionist rule extraction. In G.\ Tesauro, D.S.\ Touretzky and
T.K.\ Leen (eds.), {\it Advances in Neural Information Processing
  Systems 7}, pp.\ 609--616. Cambridge, MA: MIT Press.

[2] Bower, J.M.\ \& Beeman, D.\ (1995) {\it The Book of GENESIS:
  Exploring Realistic Neural Models with the GEneral NEural SImulation
  System.}  New York: TELOS/Springer--Verlag.

[3] Hasselmo, M.E., Schnell, E.\ \& Barkai, E.\ (1995) Dynamics of
learning and recall at excitatory recurrent synapses and cholinergic
modulation in rat hippocampal region CA3. {\it Journal of
  Neuroscience} {\bf 15}(7):5249-5262.

>>>>

% to compile a camera-ready version, add the [final] option, e.g.:
\usepackage[final]{nips_2017}
% \usepackage{nips_2017}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsmath}
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{todonotes}

%\PassOptionsToPackage{numbers}{natbib}
%\usepackage[numbers]{natbib}
%\usepackage[natbib, maxcitenames=3, style=numeric]{biblatex}
%\setcitestyle{numbers}
%\bibliographystyle{abbrvnat}

\title{Generative Modeling of Cell Images}

\author{
  Peter Goldsborough\thanks{This work has been conducted as part of an internship at the Imaging Platform of the Broad Institute.}\\
  %Imaging Platform \\
  Broad Institute of MIT and Harvard \\
  %Cambridge, MA, USA \\
  \texttt{pgoldsbo@broadinstitute.org} \\
  \And
  Nick Pawlowski \\
  %Department of Computing \\
  Imperial College London \\
  %London, UK \\
  \texttt{n.pawlowski16@imperial.ac.uk} \\
  \AND
  Juan C Caicedo \& Shantanu Singh \& Anne E Carpenter\\
  %Imaging Platform \\
  Broad Institute of MIT and Harvard \\
  %Cambridge, MA, USA \\
  \texttt{\{jccaicedo, shsingh, anne\}@broadinstitute.org} \\
}

@@

\begin{abstract}
We explore the application of Generative Adversarial Networks to the domain of morphological profiling of human cultured cells imaged by fluorescence microscopy. When evaluated for their ability to group cell images responding to treatment by chemicals of known classes, we find that adversarially learned representations are superior to autoencoder-based approaches. While currently inferior to classical computer vision and transfer learning, the adversarial framework enables useful and previously impossible interpretability of latent codes.
\end{abstract}

\section{Introduction}
Advances in high-throughput microscopy systems have enabled acquisition of large
volumes of high-resolution cell images \cite{Kraus2016}. This paves the way for
novel computational methods that can leverage these large quantities of data to
study biological systems. Our work focuses on the task of morphological
profiling, which aims to map microscopy images of cells to representation
vectors that capture salient features and axes of variation in an unsupervised
manner \cite{caicedo_profiling}. These representations ideally divide the
morphological space into clusters of cells with similar properties or, in the
case of induced perturbations, similar function.

Current techniques for morphological profiling broadly fall in two categories:
a) Classical image processing, using specialized software like CellProfiler
\cite{Carpenter2006}, to capture representations via manually-tuned segmentation
and traditional computer vision pipelines, and b) transfer learning to extract
features learned by deep convolutional neural networks originally trained to
classify miscellaneous objects \cite{pawlowski2016automating,
ando2017improving}. Classical computer vision approaches offer better
interpretability of features but require more human tuning of the segmentation
algorithms, and are limited by the feature set implemented in the image analysis
software. Current transfer learning approaches have been shown to outperform
classical methods in at least one dataset \cite{ando2017improving}. However,
given that these networks were trained on natural images (in RGB), they do not
discover the relations of the biologically meaningful image channels. Instead,
their superior performance is likely due to the their ability to capture
high-level vision features, which represent the overall structure of cells, but
not necessarily intricate details of their morphological variations. Therefore,
we hypothesized that learning representations specifically tuned to cell images
would be valuable.

We employ Generative Adversarial Networks (GANs) \cite{goodfellow2014generative}
to build a generative model of cell images from the BBBC021 dataset
\cite{Caie2010,Ljosa2012}.  We show that this generative model learns rich
feature representations and synthesizes realistic images with interpretable
properties. Our approach offers the following advantages:

\begin{itemize}
  \item \emph{Adaptability}: Unsupervised representation learning, which enables easy adaptation and incorporation of new data. This is important given that applications of morphological profiling do not have ground truth annotations, but instead aim to reveal the structure of the image collection.
  \item \emph{Specialization}: Our method learns from training data and is able to extract inherent semantic relationships. Transfer learning approaches lack this ability and cannot capture the intrinsic relationships between the biologically meaningful channels.
  \item \emph{Interpretability}: The generative abilities of this model enable previously impossible interpretbility by being able to visualize cells dependent on a given representation.
\end{itemize}

@@

Our work lies at the intersection of automated morphological profiling, and
representation learning with deep neural networks, in particular generative
architectures. Caicedo et al. \cite{caicedo2017data} recently outlined the state
and challenges of the morphological profiling problem. Prior to this, Ljosa et
al. \cite{Ljosa2013} compared the performance of various dimensionality
reduction methods for CellProfiler features on the BBBC021 benchmark
\cite{Caie2010,Ljosa2012}, consisting of MCF7 cells exposed to different
chemical treatments. Singh et al. \cite{Singh2014} demonstrated the importance
of illumination correction of those images. Pawlowski et al.
\cite{pawlowski2016automating} for the first time reported a
representation-learning method based on deep learning that is competitive with
hand-engineered features at the task of MOA prediction. Ando et al.
\cite{ando2017improving} also applied transfer learning with a different
architecture and introducing a novel feature normalization method. Further
related work on this dataset spans supervised classification \cite{Kraus2016a}
or deep learning applications based on CellProfiler features
\cite{Kandaswamy2016,Zamparo2015}

Few works have applied unsupervised deep learning techniques to the task of
feature extraction in morphological profiling. Pawlowski \cite{pawlowski2016msc}
first investigated autoencoder-based approaches but reported results far
inferior to hand-tuned features and transfer-learning approaches. Our model is
more related to the work by Osokin et al. \cite{osokin2017gans} and Johnson et
al. \cite{johnson2017generative}, wherein GANs were used to model cell
images, although their applications did not include morphological profiling.

\section{Using GANs for Interpretable Representation Learning}

Goodfellow et al. \cite{goodfellow2014generative} introduced GANs as a game of
two \emph{players}: a \emph{generator} $G$ and a \emph{discriminator} $D$. The
former receives samples $\mathbf{z}$ drawn from a noise prior $P(\mathbf{z})$
which it maps to values $G(z)$ that should resemble elements of some data
distribution $P_{\text{data}}$. The discriminator must learn to distinguish such
synthetic samples from real values $\mathbf{x} \sim
P_{\text{data}}(\mathbf{x})$. The critic's confidence in the realism of the
generator's productions is used as feedback to $G$, guiding it to synthesize
ever more realistic replicates of samples from the data prior. This procedure is
formalized in a zero-sum game, solved in a nash equilibrium found by playing the
minimax game

$$\min_G \max_D V(G, D) = \mathbb{E}_{x \sim P_{data}}[\log(D(\mathbf{x})] + \mathbb{E}_{z \sim P_{noise}}[\log(1 - D(G(\mathbf{z}))].$$

Radford et al. \cite{radford2015unsupervised} first specialized GANs to image
synthesis by introducing Deep Convolutional GANs (DCGANs). DCGANs implement the
generator and discriminator based on convolutional operations. Derivations of
DCGANs such as Least Squares GAN (LSGAN) \cite{mao2016least} or Wasserstein GAN
(WGAN) \cite{arjovsky2017wasserstein} tackle infamous instabilities in the
training procedure of early generative models. In our experiments, we found
LSGAN to be most stable, in part leading to higher quality generated images than
both DCGAN and WGAN.

The original GAN framework does not include an explicit means of performing
inference. As such, we require extensions that allow mapping of a sample
$\mathbf{x}$ drawn from the data prior to a latent representation via some
encoder transformation $E(\mathbf{x})$. A common approach is to interpret the
penultimate layer of the discriminator as latent space. The activations of this
layer serve as a source of representation vectors. These latent codes are
thought to be meaningful because the discriminator must develop a strong
internal representation of its input in order to succeed at its discrimination
task. Furthermore, this method imposes no computational overhead compared to the
vanilla GAN models.

\section{Interpretable Synthesis of Cells}\label{synthesis}

We examine if GAN models' reported ability to synthesize realistic images
transfers to the domain of cell microscopy. Our data prior consists of images
from the BBBC021 dataset of human breast cancer cells. Originally, each image
consists of three \emph{channels} corresponding to DNA, F-Actin and B-Tubulin,
cellular components visualized by fluorescence microscopy. We stack these
channels and treat them as RGB images via a simple $\text{DNA} \mapsto \text{R},
\text{$\beta$-Tubulin} \mapsto \text{G}, \text{F-Actin} \mapsto \text{B}$
mapping. We obtain ca. 1.3 million single cell images by segmenting the raw
images using CellProfiler software. Finally, we normalize the luminance of each
channel to the range [0 -- 1].

Figure \ref{fig:generated} shows examples of images generated with LSGAN, WGAN
and DCGAN architectures alongside real images. The synthetic images,
particularly those produced by LSGAN, are not only highly detailed and
realistic, but also consistent with their biological nature. For example, it is
characteristic that B-Tubulin forms a circular halo cradling the nucleus of a
cell. This property is maintained clearly in all generated images.

\begin{figure}
  \centering
  \begin{tikzpicture}
    \tikzset{frame/.style={draw, semithick, rectangle, text width=1.7cm, text height=3.5cm, rounded corners=3pt}}

  	\draw (0, 0) node {\includegraphics[width=0.6\textwidth]{generated}};

    \path (-3.3, 0) coordinate [frame] (r);
    \draw (r)+(0, -2.2) node {\textbf{Real}};

    \path (-1.07, 0) coordinate [frame] (l);
    \draw (l)+(0, -2.2) node {\textbf{LSGAN}};

    \path (1.14, 0) coordinate [frame] (w);
    \draw (w)+(0, -2.2) node {\textbf{WGAN}};

    \path (3.35, 0) coordinate [frame] (d);
    \draw (d)+(0, -2.2) node {\textbf{DCGAN}};
  \end{tikzpicture}
  \caption{Examples of real MCF7 cells from the BBBC021 dataset juxtaposed with synthetic cell images generated with LSGAN, WGAN and DCGAN. The generated images of LSGAN are most realistic and consistent with biological properties of cells.}
  \label{fig:generated}
\end{figure}

Current approaches to morphological profiling capture representations that are
often hard to interpret. While certain CellProfiler features such as nucleic
area are easy to understand, others like Zernike moments are not intuitive. For
transfer learning it is nearly impossible to visualize the captured concept. In
contrast, the GAN noise space $P(z)$ has been shown to be highly interpretable
and reveal rich semantic structure \cite{radford2015unsupervised}. We are able
to demonstrate this for images of cells. Figure \ref{fig:interpolations}
exhibits how interpolating between two noise samples $z_1, z_2 \sim P(z)$ leads
to visually smooth transitions in synthesized cells. Figure \ref{fig:algebra}
shows that $P(z)$ encodes semantic relationships, enabling algebra on
interpretable properties of the images. While there is no way to encode images
into $P(\mathbf{z})$ with the methods presented so far, we believe more advanced
models that enable this will be highly valuable if they can maintain these rich
semantic properties.

\begin{figure}
  \centering
  \includegraphics[width=\textwidth, trim={0 0 0 2.5cm}, clip]{interpolation}
  \caption{Interpolating between two points $\mathbf{z_1}, \mathbf{z_2} \sim P(\mathbf{z})$ results in visually smooth transitions in the corresponding synthesized images. Each row shows the transition from $\mathbf{z_1}$ to $\mathbf{z_2}$ from left to right.}
  \vspace{-0.1cm}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.6\textwidth]{algebra-tubulin}
  \includegraphics[width=0.6\textwidth]{algebra-nucleus}
  \caption{Vector algebra in the noise space translates to biologically valuable relationships in generated images. In the first row, subtracting the vector for a cell with small quantities of $\beta$-Tubulin (green stain) from one with high amounts yields a vector representation for "high $\beta$-Tubulin content". In the second row, the difference between a large and small nucleus encodes the semantic meaning of "larger nucleus", which can be added to vectors of other cell images to grow the size of the nucleus.}
  \label{fig:algebra}
\end{figure}

\section{Representation Learning for Morphological Profiling}\label{moa}
We test the quality of representations extracted by the discriminator used to generate single cell images learned by testing their ability to cluster treatments of similar function or mechanism-of-action. We obtain treatment profiles by averaging the extracted single cell representations found as intermediate activations of the discriminator. Further, we follow the experimental protocol of \cite{Ljosa2013} and report an average mechanism-of-action classification accuracy of a leave-one-compound-out cross-validation using a one nearest neighbor classifier.

\begin{table}
  \centering
  \begin{tabular}{lllllll}
  	\toprule
    DCGAN & WGAN & LSGAN & VAE \cite{pawlowski2016msc} & CP \cite{Singh2014} & Deep Transfer \cite{pawlowski2016automating}  & Deep Transfer \cite{ando2017improving} \\
    \midrule
    43 \%& 45 \% & 63 \% & 49 \% & 90 \% & 91 \% & 96 \% \\
    \bottomrule \\
  \end{tabular}
  \caption{Accuracy of various methods at the task of predicting the mechanism-of-action of treatments via nearest-neighbor classification in the BBBC021 dataset. CP refers to the results by Singh et al. \cite{Singh2014} using CellProfiler Features. The Deep Transfer methods correspond to Deep Feature Transfer as proposed by Pawlowski et al. \cite{pawlowski2016automating} and refined by Ando et al. \cite{ando2017improving}.}
  \label{moa-results}
\end{table}

Table \ref{moa-results} compares the accuracy of our approach to classical CellProfiler features and transfer learning. We found that the quality of representations found by unsupervised learning is not yet competitive with other established methods. Nevertheless, we believe that the adversarial approach has significant benefits regarding its interpretability as outlined above. Additionally, this framework is able to adapt to the dataset at hand and extract inherent relations that are not captured by previous methods. We believe that further gains can be made, as the image quality corresponds to classification accuracy. As such the best performing method (LSGAN) also yields the highest image quality, as judged qualitatively by expert biologists and shown in Figure \ref{fig:generated}.

@@

This work investigates the use of GANs for the domain of cell microscopy
imaging, particularly morphological profiling. First, we demonstrate that GANs
are able to synthesize biologically realistic images of cells. In addition, we
explain that GANs induce a rich semantic structure in their noise space,
providing a powerful basis for interpretability of observed properties. Second,
we add an encoding component to standard GANs and evaluate the quality of
learned representations via their mechanism-of-action classification performance
following \cite{Ljosa2013}. Even though adversarially learned representations
are currently inferior at this task, we argue that further enhancements to GANs,
especially biologically meaningful ones, will lead to higher quality latent
representations.

We emphasize that this work covers only a very small fraction of possible
applications of GANs to the domain of microscopical images. For example, we hope
that future work investigates BiGANs \cite{donahue2016adversarial} or other
novel solutions to infer latent features with GANs. We believe that improved
inference, combined with the interpretable nature of the GAN framework, may
enable simulated experiments by performing algebra with vectors corresponding to
cell lines, diseases or other perturbations. Finally, we note that deep learning
generally flourishes with larger amounts of data. As such, we conjecture that
the performance of the outlined approaches will improve with the use of larger
datasets.

\section*{Acknowledgements}
Nick Pawlowski is supported by Microsoft Research through its PhD Scholarship
Program and the EPSRC Centre for Doctoral Training in High Performance
Embedded and Distributed Systems (HiPEDS, Grant Reference EP/L016796/1).

\bibliographystyle{plain}
\bibliography{lit}
